NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-06-24 Manuscript/manuscriptPS.tex
Update text

Calcium ions form strong dehydrated molecular complexes with PS lipids,
which can also lead to phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
Such complexes and phase separation are not observed in
bilayers that are sufficiently diluted with PC
lipids \cite{feigenson86,mattai89,roux90,roux91}.
Therefore, the Ca2+ binding affinity to bilayers if often measured
from POPC:POPS (5:1) mixtures \cite{roux90,roux91}.

The headgroup order parameter changes of POPC and POPS from 
are shown in Fig. \ref{changesWITHCaClPS} as a function of Ca$^{2+}$ concentration from different
simulations and experiments \cite{roux90}. 
The ion density distributions from the simulations are shown in Fig. \ref{CAdensPCPSmixture}.
The results suggest that Ca$^{2+}$ ions clearly overbind in simulations with
MacRog and Berger models, as expected from previous results for PC lipid bilayers \cite{catte16}.
It should be noted, however, that the lowest concentration (100mM) gives
a good agreemet with experiments
\todo{Should be analyze/discuss this further?
  Binding with $\sim$100 mM is saturated in both Berger and MacRog simulations. Maybe this is realistic?
It should be noted that Berger simulation do not have counterions.}.
Surprisingly, the calcium binding seems to be too weak CHARMM36 simulations.
This is due to the NBfix interaction parameters from Ref. \citenum{kim16}, incorporated
in the parameters from CHARMM-GUI at the time of running the simulations (January 2018).
The binding of calcium to pure POPC bilayer is also too weak with these parameters
as shown in Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}.
\todo{The discussion is to be finished when we have all the data in the plot.}

\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    Changes of POPC (left) and POPS (right) headgroup order parameters in POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration. Experimental data is taken from \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero. 
  }
  \todo{Information about the cuonterions in different simulations should be added} \\
  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
  \todo{Upcoming Lipid17 simulations have been mentioned in the blog
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1515177306419\#c994825612316235467}
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{The CHARMM results are mass densities, numbers should be used.} \\
  \todo{Should we include also counterions into the plot?} \\
  \todo{Not all the data from MacRog is included.} 
\end{figure}

Also the order parameters of PS headgroup from POPC:POPS (5:1) mixture
are shown in Fig. \ref{changesWITHCaClPS} as a function of CaCl$_2$ concentration.
In experiments the order parameters exhibit a strong dependence of CaCl$_2$ with
small concentrations with a rapid saturation around 50 mM. 
The changes of PS headgroup order parameters with added CaCl$_2$ are overestimated in
all tested simulation models. Furthermore, the changes of the headgroup order parameters
do not qualitatively agree with experiments. This is in contrast to previous results
for PC headgroup \cite{catte16}, where qualitatively correct reponse to bound ions was
observed despite of significant discrepancies in the headgroup structure without additional ions.

>>>>

The dehydrated complexes of PS headgroup and calcium ions can also lead to the
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
Therefore, the Ca2+ binding affinity to PS lipids containing bilayers is easier
to study using mixtures diluted with PC lipids \cite{roux90,roux91}, where
the lipid-ion complexes and phase separation are not observed \cite{feigenson86,mattai89,roux90,roux91}.
The decrease of POPC headgroup order parameters as a function of Ca$^{2+}$ concentration in the
POPC:POPS (5:1) mixture is overestimated in all the tested simulation models
when compared with experiments \cite{roux90}, except in the CHARMM36 simulations with
special NBfix \cite{kim16} for calcium which underestimate the change (Fig. \ref{changesWITHCaClPS}).
According to the electrometer concept, this means that the calcium binding
seen in the ion density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture})
is underestimated in the CHARMM36/NBfix model, but overestimated in the other tested models.
The overbinding of Ca$^{2+}$ in simulations is expected based on previous study of
PC lipid bilayers \cite{catte16}, but underestimated calcium binding affinity in
CHARMM36/NBfix model is surprising because CHARMM36 predicted overestimated
binding to PC bilayers. The difference can be explained by the NBfix interaction
parameters from Ref. \citenum{kim16}, incorporated in the parameters give by
the CHARMM-GUI at the time of running the simulations (January 2018).
These parameters underestimate also the binding of calcium to pure POPC bilayers
(Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}).

%It should be noted, however, that the lowest concentration (100mM) gives
%a good agreemet with experiments
%\todo{Should be analyze/discuss this further?
%  Binding with $\sim$100 mM is saturated in both Berger and MacRog simulations. Maybe this is realistic?
%It should be noted that Berger simulation do not have counterions.}.

\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    Changes of POPC (left) and POPS (right) headgroup order parameters in POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration. Experimental data is taken from \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero. 
  }
  \todo{Information about the cuonterions in different simulations should be added} \\
  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
  \todo{Upcoming Lipid17 simulations have been mentioned in the blog
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1515177306419\#c994825612316235467}
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{The CHARMM results are mass densities, numbers should be used.} \\
  \todo{Should we include also counterions into the plot?} \\
  \todo{Not all the data from MacRog is included.} 
\end{figure}

The headgroup order parameters of POPS headgroup measured from POPC:POPS (5:1) mixture
exhibit a strong dependence of CaCl$_2$ with small concentrations with a rapid saturation
below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order parameter of
POPS increase with the added CaCl$_2$ in the experiment and in all the tested simulation models,
but simulations significantly overestimated the change. The larger $\alpha$-carbon order
parameter of POPS decrease and the smaller one slightly increase with the added
CaCl$_2$ in the experiment. The changes are again significantly overestimated in
the simulations, however, in this case all simulations predict qualitatively different
behaviour. Notably, the changes of POPS headgroup order parameters are overestimated
also in the CHARMM36/NBfix model where the calcium binding affinity was too low. 
We conclude that the effect of bound ions to the headgroup order parameters of POPS
is not qualitatively reproduced by the tested simulations models. This is in contrast
to previous results for PC headgroup \cite{catte16}, where qualitatively correct reponse to bound ions was
observed despite of significant discrepancies in the headgroup structure without additional ions. 
The response of POPS headgroup order parameters to the bound charge is systematic but
less well understood than the responce of PC headgroups used in the electrometer
concept \cite{seelig87,roux90}. The force field development is necessary to
generate MD simulations that could be used to explain the interactions between
PS headgroup and calcium ions.
